% Homogeneous covariance among random effects terms
% Andrey Ziyatdinov
% 14/09/2014

```{r setup, include = FALSE}
opts_chunk$set(fig.path = 'figure/', cache.path = 'cache/', echo = TRUE, 
  cache = FALSE, tidy = FALSE,
  fig.width = 9, fig.height = 9, dev = 'png', fig.cap = "",
  warning = FALSE, message = FALSE)
opts_knit$set(upload.fun = identity)
```

## About

This report aims to reproduce an example in the article 
[Fitting Linear Mixed-Effects Models using lme4](http://arxiv.org/abs/1406.5823), Appendix A.1.

The model under study has a form:

```
respVar ~ 1 + (explVar1|groupFac1) + (explVar2|groupFac2)
```

The model has two random terms, both including a random intercept and slopes.
Each random term contains a 2x2 covariance matrix.
The current version of `lme4` does not allows a model, where random terms share the same covariance matrices.
This example shows such a model thanks to the modular imlementation of the `lme4` package.

The `lme4` package version:

```{r}
library(lme4)
packageVersion("lme4")
```

## Data simulation

```{r, cache = TRUE}
set.seed(1)

nGrps <- 50

explVar <- data.frame(explVar1 = rnorm(nGrps^2), 
  explVar2 = rnorm(nGrps^2))

groupFac <- expand.grid(groupFac1 = as.factor(1:nGrps),
  groupFac2 = as.factor(1:nGrps))

randomIntercept <- expand.grid(randomIntercept1 = rnorm(nGrps),
  randomIntercept2 = rnorm(nGrps))
  
rnmdSlope <- expand.grid(randomSlope1 = rnorm(nGrps),
  randomSlope2 = rnorm(nGrps)) - randomIntercept
  
linearPredictor <- apply(randomIntercept + rnmdSlope*explVar, 1, sum)

residError <- rnorm(nGrps^2)

respVar <- linearPredictor + residError

dat <- data.frame(respVar, explVar, groupFac)

dim(dat)
head(dat)
```

## Default model

```{r, cache = TRUE}
mod1 <- lmer(respVar ~ 1 + (explVar1|groupFac1) + (explVar2|groupFac2), dat)
mod1
```

The `theta` model parameters are (expected to be of length `6`):

```
round(mod1@theta, 2)
```

The length of the vector of the random effects is (expected to be: 2 terms x 50 group levels x (slope + intercept) = `200`):

```{r}
length(mod1@u)
```

The relative covariance factor `Lambda`:

```{r}
dim(mod1@pp$Lambdat)
```

```{r}
mod1@pp$Lambdat[1:5, 1:5]
```


The number of non-zero elements in this matrix:

```{r}
length(mod1@pp$Lind)
```




